---
title: "Cleaning and Cropping Quadrat Data"
description: >
  A complex walk through of how to clean up quadrat data using the `quadcleanR`
  package. This will go through all the functions inclusing the `croparea()`
  function, as well as visualizing the resulting data. This vignette is good
  for those who have to crop quadrats and clean up the data. 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cleaning and Cropping Quadrat Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette will walk you through cleaning up quadrat data to produce easy to analyze data frames.

A little about the data:

The data being cleaned will be the `softcoral_SQuads` and this vignette will build off of the cropping vignette as well. Data was collected by the Baum Lab and Kiritimati Field Teams and is the uncleaned version of data found in Maucieri and Baum 2021. Biological Conservation. The `softcoral_SQuads` data are from photo quadrats (0.9m by 0.6m) which were randomly annotated with 54 random points each, while the `softcoral_annotations` data are from photo quadrats (1m by 1m) which were randomly annotated with 100 random points each. At each of these annotated points, the substrate was identified. Photo quadrats were collected on Kiritimati Island in the Republic of Kiribati and document coral cover over time and space. The annotations and output of the data frame were produced using [CoralNet](https://coralnet.ucsd.edu/) and all annotations were done manually, by trained researchers.

So this vignette will follow the following steps:

1. Take the small quadrats (`softcoral_SQuads`) and show how to clean the data

2. Then crop the large quadrat data (`softcoral_annotations`) and clean it too

3. Join the two together to produce a large, easy to use data frame for future analyses, with some further cleaning of this joint data set. 

First lets load the package, the dplyr package, tidyr package and the data

```{r load package, message = FALSE, warning = FALSE}

library(quadcleanR)
library(dplyr)
library(tidyr)
library(gtools)

data("softcoral_SQuads")

```


Now let me point out some unique aspects of this data:

```{r head and tail, message = FALSE}

head(softcoral_SQuads)
tail(softcoral_SQuads)

```
This data as an `Image.ID` column which was arbitrarily added to this data set, so we are going to remove that as it holds no scientific value. There is also a final row which sums all quadrats but since we will be removing quadrats and points to clean these data up, we will remove that final row as well. The `Image.name` column is the unique ID for each photo quadrat, but it is very messy and not easy to use, so we will make this into new columns and add more information. `Annotation.status` is a column from [CoralNet](https://coralnet.ucsd.edu/) which tells if the annotations in each photo quadrat have been confirmed by human researchers or are only based on AI. The `Points` column tells us how many randomly annotated points there are in each quadrat, and since they are all 54, we know these data are from the smaller quadrats. The rest of the columns are the different coral and substrate IDs and how many points where annotated for each tag in each photo.

So first, lets remove unneeded columns and make sure we are only working with "Confirmed" annotations.

```{r confirmed annotation status}

SQuad_confirmed <- softcoral_SQuads %>% filter(Annotation.status == "Confirmed") %>% select(-c(Image.ID, Points, Annotation.status))

```

Now we will separate the `Image.name` column into more descriptive columns.

```{r separating ID, warning=FALSE}

SQuad_separated <- separate(SQuad_confirmed, Image.name, sep="_", into=c("Field.Season", "Site","Quadrat"))

```

There will be 3 rows with discarded pieces because the quadrats have an _ in them, but that is okay, they will still be unique quadrat names so we don't need to worry about that. But if you notice, there are still .jpg and .jpeg in the quadrat names, so lets remove those, and change the naming of siteT19 to site40

```{r remove jpg and T19}

SQuad_nojpg <- rm_chr(SQuad_separated, c(".jpg", ".jpeg"))
SQuad_site40 <- change_values(SQuad_nojpg, "Site", "siteT19", "site40")

```

Now lets look at the levels of some of my columns. 

```{r view levels}

unique(SQuad_site40$Field.Season)
unique(SQuad_site40$Site)

```


Sometimes you may not want every year and every site of the data, so you may want to subset some of them out. For this example I don't want to subset any of these out, but I will demonstrate how to subset out based on values later, when we get to the large quadrats. 


Now to prep for the addition of the large quadrats, I want to update the column names for this data frame. The column names are currently set as the tag shorthand used during the annotation process, but now I want them to better reflect the actual substrate names.

```{r add labelset, message=FALSE}

data("coral_labelset")
head(coral_labelset)

```

This is what my label set document looks like, but you could also make this in R by joining a series of vectors. Now lets fix the column names.

```{r change colnames}

SQuad_colnames <- change_names(SQuad_site40, coral_labelset, "short_name", "full_name")
names(SQuad_colnames)

```

Much better.

Now because these are smaller quadrats with fewer annotated points, before we add in the large quadrat data, lets deal with the substrate that was not able to be accurately identified. For these photo quadrats, the Shadow, Transect_hardware and Unclear tags need to be removed and not used when we calculate percent cover. If I was going to use this data to do a diversity analysis with hard corals, I would also include unknown_hard_coral, and Bleached_unknown_hard_coral to this list, but as this data will be used for soft coral analyses, we will leave it. 

```{r usable points}

SQuad_colnames <-  mutate_at(SQuad_colnames, c(4:134), as.numeric)

SQuad_usable <- useable_obs(SQuad_colnames, c("Shadow", "Transect_hardware", "Unclear"), 
                            max = TRUE, cutoff = 10)

SQuad_removed <- useable_obs(SQuad_colnames, c("Shadow", "Transect_hardware", "Unclear"),
                             max = TRUE, cutoff = 10, print_max = TRUE)

```

By identifying how many useable points there are in each quadrat, and removing any quadrats that had over 10% of the identified points unuseable, we have removed 0 quadrats from analysis, which you could view with the `SQuad_removed` data frame but as we have no quadrats to removed, it will be an empty data frame.

Now we can start working on the large quadrat data. 

```{r large quad}

data("softcoral_annotations")
data("softcoral_LQuads")

```


I will be cropping this data as outlined in the cropping vignette, so check that out for the details, but before I crop it, I want to remove any unconfirmed quadrats just as we did for the small quads.

```{r removing unconfirmed}

LQuad_confirmed <- softcoral_LQuads %>% filter(Annotation.status == "Confirmed") 

LQuad_sub <- softcoral_annotations %>% filter(Name %in% unique(LQuad_confirmed$Image.name))

```

Then we will crop the annotations data. 

```{r large quad cropping}

LQuad_cropped <- crop_area(data = LQuad_sub, row = "Row",
                             column = "Column", id = "Name", dim = c(0.9, 0.6), 
                             obs_range = c(36,64))
```

Now I will just format this to look like the small quad data so we can join the two.

```{r large quad formatting}

LQuad_wide <- LQuad_cropped %>% select(-c(obs, Row, Column)) %>% group_by(Name) %>% pivot_wider(names_from = Label, values_from = Label, values_fn = length, values_fill = 0)
LQuad_wide <- as.data.frame(LQuad_wide)

LQuad_separated <- separate(LQuad_wide, Name, sep="_", into=c("Field.Season", "Site","Quadrat"))

LQuad_nojpg <- rm_chr(LQuad_separated, c(".jpg", ".jpeg"))

LQuad_colnames <- change_names(LQuad_nojpg, coral_labelset, "short_name", "full_name")


LQuad_colnames <-  mutate_at(LQuad_colnames, c(4:11), as.numeric)

LQuad_usable <- useable_obs(LQuad_colnames, c("Shadow", "Transect_hardware", "Unclear"),
                             max = TRUE, cutoff = 0.1*54)

LQuad_removed <- useable_obs(LQuad_colnames, c("Shadow", "Transect_hardware", "Unclear"),
                             max = TRUE, cutoff = 0.1*54, print_max = TRUE)

```

With the large quadrats, only 4 photos needed to be removed because they had too many unuseable points. 

There are a few more things that need to be removed from the large quadrat data, including MPQs (Mega photo quadrats), site8.5 and DEEP photo quadrats. 

```{r remove DEEP, site 8.5 and MPQs}

LQuad_noDEEP_site8.5 <- keep_rm(LQuad_usable, c("DEEP", "site8.5"), select = "row", exact = FALSE, colname = "Site", keep = FALSE)

LQuad_noMPQ <- keep_rm(LQuad_noDEEP_site8.5, c("MPQ"), select = "row", exact = FALSE, colname = "Quadrat", keep = FALSE)

```

Alright, not we can join these data frames together, and continue with one large data set. As both data sets used the samle label set of names, This is easy with the `smartbind()` function out of the `gtools` package

```{r bind quad dfs}

# AllQuads <- smartbind(SQuad_usable, LQuad_noMPQ, fill = 0) %>% select(-c(unusable))

```


Now we have removed unusable annotations, so lets rescale and recalculate cover with proportion cover. This data frame has ~6500 rows so this will take a minute or two.

```{r prop cover calc}

# AllQuads_cover <- cover_calc(AllQuads, names(AllQuads[,4:131]), prop = TRUE)

```
 
One thing you may notice by looking at the column names, is that these columns are not specific species, but there are duplicates of the same species, categorized into bleaching and non bleaching forms. For any kind of diversity analysis, this would inflate the number of different species, so it is important to combine different forms of the same species if diversity analyses are being done.

This data frame is now nicely formatted and could be used for many community based analyses. This might be a great stopping point for some analyses, but to further clean this up I am going to convert this into long format data.

```{r}

# AllQuads_long

```













